```
---
title: "OmniCorr: Salmon multi-omics integration workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OmniCorr – Salmon Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates the **core OmniCorr workflow** using the salmon multi-omics dataset  
(which serves as the main example in the
OmniCorr manuscript).

Main steps shown in this document:

- Loading processed example data  
- Checking & fixing sample alignment  
- Computing cross-omics correlations  
- Visualizing results using aligned heatmaps

All preprocessing steps (from raw sequencing → processed data)  
are provided separately in the `raw-data/` directory of the repository.

### Data provenance & preprocessing

The example datasets distributed with **OmniCorr** were generated from:

- Transcriptomics  
- Metagenomics  
- Metatranscriptomics

Complete preprocessing workflows (normalization, filtering,
WGCNA-based feature clustering, etc.) are documented in:

```
raw-data/
├── transcriptomics/
│   └── Transcriptomics_preprocessing.Rmd
├── metagenomics/
│   └── Metagenomics_preprocessing.Rmd
└── metatranscriptomics/
    └── Metatranscriptomics_preprocessing.Rmd
```

The final processed objects are saved as `.rda` files and shipped with the package.

## Load OmniCorr + example data

```{r load-data}
library(OmniCorr)

data(Transcriptomics)
data(Metagenomics)
data(Metatranscriptomics)
data(metadata)
```

**Each dataset contains:**

- **rows** = samples  
- **columns** = features / modules / hubs / taxa / ...

## Very important: Sample alignment (required!)

All omics layers **must** contain exactly the same samples in the **same order**.

```{r check-sample-alignment}
# Quick check
all(
  rownames(Transcriptomics) == rownames(Metagenomics) &
  rownames(Transcriptomics) == rownames(Metatranscriptomics)
)
```

If order is wrong / samples are missing → use the helper function:

```{r fix-sample-order}
df_list <- CheckSampleOrder(Transcriptomics, Metagenomics)

Transcriptomics <- df_list$df1
Metagenomics <- df_list$df2

df_list <- CheckSampleOrder(Transcriptomics, Metatranscriptomics)

Transcriptomics <- df_list$df1
Metatranscriptomics <- df_list$df2
```

## Workflow steps

### Step 1 – Feature clustering (using reference layer = Transcriptomics)

```{r feature-clustering}
library(WGCNA)

dendro <- hclust(
  d = as.dist(1 - bicor(Transcriptomics, maxPOutliers = 0.05)),
  method = "ward.D2"
)
```

### Step 2 – Transcriptomics reference heatmap

```{r reference-heatmap}
library(pheatmap)

tx_heatmap <- pheatmap(
  t(Transcriptomics),
  cluster_rows = dendro,
  cluster_cols  = FALSE,
  show_rownames = FALSE,
  main          = "Transcriptomics"
)
```

→ This heatmap will serve as **structural backbone** for all following plots

### Step 3 – Cross-omics correlation analysis

```{r cross-omics-correlations}
# Transcriptomics ↔ Metagenomics
corr_mg <- calculate_correlations(
  df1 = Transcriptomics,
  df2 = Metagenomics,
  show_significance = "stars"
)

# Transcriptomics ↔ Metatranscriptomics
corr_mt <- calculate_correlations(
  df1 = Transcriptomics,
  df2 = Metatranscriptomics,
  show_significance = "stars"
)
```

Each result contains:

- `$correlation`
- `$p_value`
- `$p_value_adj`
- `$signif_matrix` (the stars/symbols matrix)

### Step 4 – Beautiful aligned correlation heatmaps

```{r correlation-heatmaps}
library(RColorBrewer)

heatmap_colors <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(51)


# Metagenomics correlations
hm_mg <- pheatmap(
  corr_mg$correlation,
  color         = heatmap_colors,
  cluster_rows  = dendro,
  display_numbers = corr_mg$signif_matrix,
  breaks        = seq(-1, 1, length.out = 51),
  show_rownames = FALSE,
  legend        = FALSE,
  main          = "Metagenomics"
)


# Metatranscriptomics correlations
hm_mt <- pheatmap(
  corr_mt$correlation,
  color         = heatmap_colors,
  cluster_rows  = dendro,
  display_numbers = corr_mt$signif_matrix,
  breaks        = seq(-1, 1, length.out = 51),
  show_rownames = TRUE,
  legend        = TRUE,
  main          = "Metatranscriptomics"
)
```

### Nice combined layout (optional but very recommended)

```{r combined-layout}
library(cowplot)

plot_grid(
  tx_heatmap$gtable,
  hm_mg$gtable,
  hm_mt$gtable,
  ncol       = 3,
  align      = "h",
  rel_widths = c(3, 1.3, 2.2)
)
```

### Bonus: Correlation with metadata

```{r metadata-correlation}
corr_meta <- calculate_correlations(
  df1 = Transcriptomics,
  df2 = metadata,
  use = "pairwise.complete.obs",
  show_significance = "stars"
)

pheatmap(
  corr_meta$correlation,
  color         = heatmap_colors,
  cluster_rows  = dendro,
  display_numbers = corr_meta$signif_matrix,
  breaks        = seq(-1, 1, length.out = 51),
  show_rownames = TRUE,
  main          = "Environmental variables / Metadata"
)
```

## Summary

This vignette shows the **core OmniCorr workflow** using the salmon multi-omics example.

**Where to look next:**

- Full preprocessing → `raw-data/` folder  
- All processed example data → inside the package (`data()`)
- Other case studies → look for Case_Study_2 (cattle)

Good luck with your multi-omics integration projects! 🐟

```{r session-info}
sessionInfo()
```
